### 2020-03-26 Paper_shortened_pdf.Rmd
Improve discussion

# Discussion

This study set out to obtain a more comprehensive and nuanced understanding of how fathers' and mothers' quantitative involvement relate to children's emotional adjustment development. We used innovative methods that control for stable between-family differences, in order to obtain more accurate estimates of within-family effects and investigate the direction of effects between parents and children. Furthermore, we examined how joint and relative quantitative involvement predict parenting and emotional adjustment, and moderate the longitudinal interplay between qualitative parenting and emotional adjustment.

## Associations between qualitative parenting and emotional adjustment

We found mixed evidence, at both the within- and between-family levels, for our hypotheses that warmth, consistency, and inductive reasoning would be positively associated, and anger would be negatively associated, with emotional adjustment. Parental warmth was not significantly correlated with adjustment at the between-family level, but changes in warmth were correlated with changes in adjustment at the within-family level. Consistency was positively correlated with adjustment for both parents at the between-family level, indicating that children in families with highly consistent parents tended to be more emotionally well-adjusted, and for mothers at the within-level. Counter to hypotheses, inductive reasoning was negatively correlated with adjustment at the between-family level, indicating that parents who engage in more inductive reasoning had children with poorer emotional adjustment. At the within-family level, increases in fathers' inductive reasoning were positively correlated with adjustment. Finally, for anger, we found the predicted negative associations at the between-family level only, which means that, in families where mothers or fathers display high anger, children tend to be less emotionally well-adjusted.

The only parenting effect that was significant on average, was for maternal warmth, which positively predicted emotional adjustment in most families. All within-family predictive effects varied significantly across families, however, which highlights the importance of capturing and explaining these differences. This variability was notably greater for fathers than mothers, which suggests that fathers' influence is more context-dependent.

Regarding child effects, as hypothesized, well-adjusted children evoked greater warmth from both parents on average. Counter to predictions, we did not find child effects for consistency. In partial support of our hypothesis, we found that children who displayed better emotional adjustment in one wave evoked greater inductive reasoning from mothers in the next wave. But paradoxically, and counter to our hypothesis, we also found that children who displayed better emotional adjustment in one wave evoked greater maternal anger in the next wave. Perhaps the two go hand in hand: When mothers get angry, they (eventually) also explain what provoked their anger, through inductive reasoning.

@@

We found ample evidence for within-family predictive effects between parental warmth and emotional adjustment, which were bi-directional for mothers. This suggests that warmth and emotional adjustment reciprocally influence one another within families; potentially through causal mechanisms. For consistency, however, we found no within-family predictive effects, but only between-family level associations. This suggests that the benefits for emotional adjustment associated with greater consistency might be primarily due to stable between-family differences. For inductive reasoning, paradoxically, between-family differences and within-family fluctuations showed opposing patterns of associations with adjustment. One potential explanation for this finding might be that the construct of inductive reasoning confounds *how* parents discipline their children (sample item: "How often do you talk it over and reason with this child..."), with *how often* their child misbehaves ("...when he/she misbehaved?"). The fact that between-family correlations between inductive reasoning and emotional adjustment were negative suggests that between-family differences in inductive reasoning primarily capture variance related to the frequency of misbehavior. The within-family fluctuations, by contrast, might tap into changes in how parents discipline their children to a greater extent.

For anger, we found moderately strong between-family level correlations for both parents. Counter to our expectations, however, we found no evidence for within-family correlations or parenting effects. This suggests that the brunt of negative associations between anger and adjustment is explained by between-family differences. One potential explanation for the absence of within-family associations might be a ceiling effect: If the stable level of parental anger in a family is high enough to negatively impact children's emotional adjustment, over-time fluctuations of that anger might no longer make much of a difference. We also found a paradoxical positive predictive effect of emotional adjustment on maternal anger. This means that, if children demonstrated better emotional adjustment than usual in one wave, their mothers tended to display more anger than usual in the next wave. A possible explanation might be regression to the mean: On average, emotional adjustment showed no-to-negative stability coefficients in all models. That means that better emotional adjustment in one wave, was likely to be followed worse adjustment in the next wave. The literature on expectation-based loss aversion suggests that mothers might be more angered by a child who unexpectedly displays poor adjustment, than they would be appeased by a child who displays unexpected positive adjustment [@koszegiModelReferenceDependentPreferences2006].

## Differences between fathers and mothers

Overall, we found somewhat more evidence for associations of emotional adjustment with mothers' qualitative parenting, than with fathers' parenting, and we found some evidence for differences between mothers' and fathers' roles in emotional adjustment development. These differences all played out at the within-family level, however. Stable between-family differences in either parent's qualitative parenting were all similarly associated with children's adjustment.

At the within-family level, associations with emotional adjustment were stronger for maternal consistency, and for paternal reasoning and warmth. As mentioned before, mothers tend to spend more time with their children than fathers; also in our sample. This might help explain the stronger associations for maternal consistency - a behavior which implies, by definition, consistent repeated interactions with the child. For fathers, associations were stronger for inductive reasoning and warmth - behaviors which can also be expressed in one-off interactions, such as when  disciplining a child (inductive reasoning), or during play time (warmth). The stronger association with inductive reasoning for fathers is in line with the classic view of the father as disciplinarian [@flouriFathersBehaviorsChildren2010].

With regard to parenting effects, the only difference between mothers and fathers was for the predictive effect of warmth on emotional adjustment, which was evident only for mothers. This is in line with prior work, indicating that warmth is an important socialization mechanism for mothers [@vanlissaRoleFathersMothers2018a]. With regard to child effects, we found substantial evidence that mothers' parenting was more responsive to children's emotional adjustment than fathers' parenting. Specifically, mothers' anger, inductive reasoning, and warmth were predicted by adjustment, whereas these effects were absent for fathers.

## Quantitative involvement and qualitative parenting behaviors

Turning to our findings from linkages between quantitative involvement and qualitative parenting behaviors, our results were in line with our hypothesis that greater involvement would be associated with more desirable parenting behaviors [@sabattiniRelationMothersFathers2004]. Our results revealed that, when both parents were more involved, they displayed greater warmth, consistency, and inductive reasoning. Fathers, in particular, reported substantially greater parental warmth when both parents were more involved. We found that, when mothers performed more care tasks than fathers, mothers displayed more warmth, and fathers displayed less warmth, consistency, and inductive reasoning. This increase in maternal warmth might reflect an attempt to compensate for fathers' absence [@shinnFatherAbsenceChildren1978]. These results indicate that, when childcare is unequally distributed, fathers engage in less desirable qualitative parenting behaviors [see also @craigHowMothersFathers2011].

@@

Our operationalization of quantitative involvement, in terms of joint involvement and relative involvement, allowed us to addresses the claim that any effects of father involvement on children's adjustment might in fact be due to children receiving greater total care [@deutschPaternalParticipationChild2001]. However, counter to our hypothesis, joint involvement was *negatively* associated with between-family differences in emotional adjustment. One potential explanation is suggested by self-determination theory: Excessive levels of involvement might undermine children's autonomy needs, thereby leading to increased emotional maladjustment [@vandergiessenPerceivedAutonomySupport2014]. Research on "helicopter parenting" supports this interpretation, as psychological well-being was found to be lower in children of overly involved parents [@schiffrinHelpingHoveringEffects2014]. Another potential explanation is that the amount of quantitative care provided is not only driven by parents' motivation and ability to be involved in their children's lives, but also by children's objective needs. Factors related to children's objective needs - such as developmental delay, or atypical development - might drive an increase in parents' quantitative involvement, and might also be associated with children's emotional maladjustment, resulting in a negative correlation between the two.

@@

## Interactions between quantitative and qualitative involvement

We found limited evidence that quantitative involvement moderates within-family parenting effects. We did find both joint- and relative quantitative involvement moderated the effect of maternal warmth on adjustment. Mothers' warmth predicted emotional adjustment less strongly when both parents were more involved, and more strongly when mothers were relatively more involved than fathers. These findings suggests a compensatory effect, whereby mothers' warmth can make up for disadvantages incurred by having a relatively uninvolved father [@shinnFatherAbsenceChildren1978].

Evidence that quantitative moderates child effects was somewhat stronger. Firstly, when joint involvement was higher, and when mothers' relative involvement was greater compared to fathers, emotional adjustment became less predictive of maternal warmth. These findings can be interpreted in the context of the sizable average positive effect of adjustment on maternal warmth. This positive effect was somewhat smaller in families where mothers' involvement was high - either in terms of joint, or relative involvement. Secondly, joint quantitative involvement positively moderated the link from emotional adjustment to maternal inductive reasoning. This suggests that mothers have more of a role as disciplinarian in families where both parents are highly involved. When mothers were relatively more involved than fathers, fathers decreased their inductive reasoning more strongly in response to children's positive adjustment, and increased it more strongly in response to maladjustment. Together, these findings suggests that fathers take more of a role as disciplinarian when mothers are relatively more involved in care tasks.

@@

In line with prior research, we found that both parents displayed more anger and inductive reasoning, and less warmth, towards sons than towards daughters. These findings suggest that parents are stricter with sons. Sex did not moderate parenting effects, which suggests that boys and girls are similarly affected by qualitative parenting behaviors. Sex did moderate child effects, however: Emotional adjustment predicted both parents' inductive reasoning more negatively when children were male. This finding suggests that some of parents' greater strictness toward boys is alleviated when boys are well-adjusted.

## Summary

To recapitulate, we set out to examine the direction of within-family effects by controlling for stable between-family differences. We found only limited evidence for within-family predictive effects, most of which were child effects on parents. Instead, we mostly found evidence for between-family level correlations. We are not the first to find that parent effects are less prevalent once stable between-family differences are accounted for [@keijsersParentalMonitoringAdolescent2016]. In the area of parenting and emotional adjustment development, we are aware of only one prior study that controlled for stable between-family differences, and similarly reported mostly child effects and between-family level correlations in a sample of adolescents [@vanlissaRoleFathersMothers2018a]. The authors explained their findings by proposing that child effects might be more common in adolescence than in childhood, because parent-adolescent relationships are more egalitarian at that age. They argued that the between-family level correlations might reflect the consolidated influence of within-family parent effects that occurred at an earlier age. The fact that the present study similarly found a preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that either parent effects occurred at an even younger age, or that something else is at play.

@@

Future research might also devote greater attention to between-family level correlations as an outcome of interest. These correlations can reflect the consolidated effects of parenting effects that occur at an early age [@vanlissaRoleFathersMothers2018a], or alternatively, they can reflect the influence of third variables that impact both constructs. We should seek to understand why between-family level correlations are so prominent; for example, by comparing their strength in samples with children of different ages, including very young children, or by examining the influence of more sociological factors. For example, economic hardship undermines parents' capacity for consistent parenting, as well as children's socio-emotional adjustment [@mcloydImpactEconomicHardship1990]. Due to disciplinary differences, such sociological factors are rarely examined in conjunction with sophisticated models of reciprocal influences between parents and children.

>>>>

# Discussion
<!-- To recapitulate, we set out to investigate how fathers' and mothers' quantitative and qualitative involvement relate to children's emotional adjustment development. We used innovative methods to disentangle stable between-family differences from within-family fluctuations. 
-->
This study set out to obtain a more comprehensive and nuanced understanding of how fathers' and mothers' quantitative and qualitative involvement relate to children's emotional adjustment development. We used innovative methods to disentangle stable between-family differences from within-family fluctuations in the study variables, allowing us to investigate the direction of effects between parents and children. Furthermore, we examined how joint and relative quantitative involvement predict parenting and emotional adjustment, and moderate the longitudinal interplay between qualitative parenting and emotional adjustment.

## Correlations between qualitative parenting and emotional adjustment

We found mixed evidence for our hypotheses that warmth, consistency, and inductive reasoning would correlate positively, and anger negatively, with emotional adjustment. Parental warmth was only correlated with emotional adjustment at the within-family level. Consistency was positively correlated with adjustment at the between-family level for both parents, and at the within-family level for mothers. Correlations with inductive reasoning were, paradoxically, opposite in sign at the between- and within-family levels. At the between-family level, parents who engaged in more inductive reasoning had children with poorer emotional adjustment. At the within-family level, fathers' inductive reasoning was positively correlated with adjustment. Finally, for anger, we found the predicted negative associations at the between-family level only, which means that, in families where mothers or fathers display high anger, children tend to be less emotionally well-adjusted. Although most of these findings are in line with hypotheses based on prior research that did not disentangle between- and within-family variance, the few deviations from the expected pattern illustrate an important insight: Correlations do not necessarily occur at both of these levels, and can even be opposite in sign; a phenomenon known as the ecological fallacy [see @keijsersParentalMonitoringAdolescent2016].

A potential explanation for the paradoxical findings for inductive reasoning might be that the construct of inductive reasoning confounds *how* parents discipline their children (sample item: "How often do you talk it over and reason with this child..."), with *how often* their child misbehaves ("...when he/she misbehaved?"). Stable between-family differences in inductive reasoning might tap into *how often* the child misbehaves. The negative between-family correlations between inductive reasoning and emotional adjustment suggest that poorly adjusted children are disciplined more often. Within-family fluctuations, by contrast, might tap into context-specific changes in *how* parents discipline their children to a greater extent [@critchleyParentDisciplinaryBehavior2006]. When parents use more inductive reasoning than they usually do, talking over and reasoning about misbehavior, this might benefit emotional adjustment.

@@

In terms of the direction of within-family predictive effects, we found ample evidence for within-family predictive effects between parental warmth and emotional adjustment, which were bi-directional for mothers. This suggests that parental warmth and children's emotional adjustment reciprocally influence one another within families; potentially through causal mechanisms. There was no other evidence for parenting effects however. Child effects outweighed parenting effects both in number and in size. The direction of effects thus appears to flow primarily from children to parents.

With regard to these child effects, on average, children who displayed better emotional adjustment in one wave evoked more warmth from both parents, and more inductive reasoning from mothers in the next wave. This latter finding is in line with the aforementioned within-family correlations for inductive reasoning. Counter to our hypothesis, we found that children who displayed better emotional adjustment in one wave evoked greater maternal anger in the next wave. One possible explanation is regression to the mean: On average, emotional adjustment showed no-to-negative stability coefficients in all models. That means that better emotional adjustment in one wave, was likely to be followed worse adjustment in the next wave. Looking beyond direction of effects, it is worth noting that the brunt of negative associations between anger and adjustment appears to be due to stable between-family differences. The sparsity of within-family effects of anger might be due to a ceiling effect: If the stable level of parental anger in a family is high enough to negatively impact children's emotional adjustment, over-time fluctuations of that anger might no longer make much of a difference.

## Differences between mothers and fathers
<!-- All within-family predictive effects varied significantly across families, however, which highlights the importance of capturing and explaining these differences. This variability was notably greater for fathers than mothers, which suggests that fathers' influence is more context-dependent. -->

Overall, we found somewhat more evidence for associations of emotional adjustment with mothers' qualitative parenting, than with fathers' parenting, and we found some evidence for differences between mothers' and fathers' roles in emotional adjustment development at the within-family level. These findings are in line with existing theory [@faganShouldResearchersConceptualize2014; @paquetteTheorizingFatherchildRelationship2004]. Stable between-family differences in mothers' and fathers' parenting behaviors were all similarly associated with children's adjustment, however.

Within-family associations of emotional adjustment with consistency were stronger for mothers, and associations with warmth and reasoning were stronger for fathers. As mentioned before, mothers tend to spend more time with their children than fathers; also in our sample. This might explain why associations for maternal consistency were stronger: Spending more time with the child provides greater opportunity for consistent behavior. For fathers, associations were stronger for inductive reasoning and warmth - behaviors which can also be expressed in one-off interactions, such as when  disciplining a child (inductive reasoning), or during play time (warmth). The stronger association with inductive reasoning for fathers is in line with the classic view of the father as disciplinarian [@flouriFathersBehaviorsChildren2010].

With regard to parenting effects, the only difference between mothers and fathers was for the predictive effect of warmth on emotional adjustment, which was evident only for mothers. This is in line with prior work, indicating that warmth is an important socialization mechanism for mothers [@vanlissaRoleFathersMothers2018a]. With regard to child effects, we found substantial evidence that mothers' parenting was more responsive to children's emotional adjustment than fathers' parenting. Specifically, mothers' anger, inductive reasoning, and warmth were predicted by adjustment, whereas these effects were absent for fathers. This is in line with some prior studies, which found that mothers are more sensitive to children's needs [see @faganShouldResearchersConceptualize2014, p. 395].

## Does quantitative involvement predict qualitative parenting behaviors?

Turning to our findings from linkages between quantitative involvement and qualitative parenting behaviors, our results were in line with our hypothesis that greater involvement would be associated with more desirable parenting behaviors [@sabattiniRelationMothersFathers2004]. Our results revealed that, when both parents were more involved, they displayed greater warmth, consistency, and inductive reasoning. Fathers, in particular, reported substantially greater parental warmth when both parents were more involved. We also found that, when mothers performed more care tasks than fathers, mothers displayed more warmth, and fathers displayed less warmth, consistency, and inductive reasoning. This increase in maternal warmth might reflect an attempt to compensate for fathers' absence [@shinnFatherAbsenceChildren1978]. These results indicate that, when childcare is unequally distributed, fathers engage in less desirable qualitative parenting behaviors [see also @craigHowMothersFathers2011].

@@

Our operationalization of quantitative involvement, in terms of joint involvement and relative involvement, allowed us to addresses the claim that any effects of father involvement on children's adjustment might in fact be due to children receiving greater total care [@deutschPaternalParticipationChild2001]. However, counter to our hypothesis, joint involvement was *negatively* associated with between-family differences in emotional adjustment. One potential explanation is suggested by self-determination theory: Excessive levels of involvement might undermine children's autonomy needs, thereby leading to increased emotional maladjustment [@vandergiessenPerceivedAutonomySupport2014]. Research on "helicopter parenting" supports this interpretation, as psychological well-being was found to be lower in children of overly involved parents [@schiffrinHelpingHoveringEffects2014]. Another potential explanation is that the amount of quantitative care provided is driven, in part, by children's objective needs, such as atypical or delayed development. Such factors require increased quantitative care, and are also associated with children's emotional maladjustment, which could result in a negative correlation between the two.

@@

## Does quantitative involvement moderate links between qualitative parenting and emotional adjustment?

We found limited evidence that quantitative involvement moderates parenting effects on emotional adjustment. Specifically, joint- and relative quantitative involvement moderated the effect of maternal warmth on adjustment: Mothers' warmth predicted emotional adjustment less strongly when both parents were more involved, and more strongly when mothers were relatively more involved than fathers. These findings suggests a compensatory effect, whereby mothers' warmth can make up for disadvantages incurred by having a relatively uninvolved father [@shinnFatherAbsenceChildren1978].

Evidence that quantitative involvement moderates child effects was somewhat stronger. Firstly, when joint involvement was higher, and when mothers' relative involvement was greater compared to fathers, emotional adjustment became less predictive of maternal warmth. These findings can be interpreted in the context of the sizable average positive effect of adjustment on maternal warmth. This positive effect was somewhat smaller in families where mothers' involvement was high - either in terms of joint, or relative involvement. Secondly, joint quantitative involvement positively moderated the link from emotional adjustment to maternal inductive reasoning. This suggests that mothers have more of a role as disciplinarian in families where both parents are highly involved. When mothers were relatively more involved than fathers, fathers decreased their inductive reasoning more strongly in response to children's positive adjustment, and increased it more strongly in response to maladjustment. Together, these findings suggests that fathers take more of a role as disciplinarian when mothers are relatively more involved in care tasks.

@@

In line with prior research, we found indications that parents are stricter with sons than with daughters; specifically, parents displayed more anger and inductive reasoning, and less warmth, towards sons. Sex also moderated child effects: Emotional adjustment predicted both parents' inductive reasoning more negatively when children were male. This finding suggests that some of parents' greater strictness toward boys is alleviated when boys are well-adjusted.

## Theoretical implications of disentangling within- and between-family variance

The present study mostly found evidence for correlations at the between-family level; evidence for within-family predictive effects was limited, and revealed mostly child effects. In a previous study with a comparable design, conducted in an adolescent sample, we similarly found mostly between-family level correlations and child effects [@vanlissaRoleFathersMothers2018a; see also @keijsersParentalMonitoringAdolescent2016]. We explained that finding in light of the relative equality of parent-adolescent relationships. We theorized that the between-family level correlations might reflect the consolidated influence of within-family parent effects that took place at an earlier age, and hypothesized that parenting effects would be more prevalent, and between-family correlations weaker, in younger children. The present findings refute these conjectures. The preponderance of child effects and between-family level correlations in a sample of younger children, age 6-12, suggests that either parent effects occurred at an even younger age, or that something else is at play.

@@

Future research might also devote greater attention to between-family level correlations as an outcome of interest. These correlations can reflect the consolidated effects of parenting effects that occur at an early age [@vanlissaRoleFathersMothers2018a], or alternatively, they can reflect the influence of third variables that impact both constructs. We should seek to understand why between-family level correlations are so prominent by comparing between-family correlations across children of different ages, including very young children, and by examining the influence of between-family covariates, including sociological factors. For example, economic hardship undermines parents' capacity for consistent parenting, as well as children's socio-emotional adjustment [@mcloydImpactEconomicHardship1990]. Due to disciplinary differences, such sociological factors are rarely examined in conjunction with sophisticated models of reciprocal influences between parents and children.
